---
title: "Assignment 1 - Survival Analysis (STAT 559) "
author: "Naresh Neupane"
date: "January 26, 2024"
output:
  html_document:
    df_print: paged
    toc: no
     
  word_document:
    toc: yes
  pdf_document:
    keep_tex: true
   
    mainfont: "Computer Modern"
---
<style type="text/css">

h1.title {
  font-size: 38px;
  color: DarkRed;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
  text-align: center;
}

body{
  font-family: Computer Modern;
  font-size: 12pt;
}

/* Headers */
h3,h5,h6{
  font-size: 24pt;
}

.math {
  font-size: small;
}
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

The lifetime of light bulbs follows an exponential distribution with a hazard rate of 0.001 failures per hour of use.

(a) Find the mean lifetime of a randomly selected light bulb.
(b) Find the median lifetime of a randomly selected light bulb.
(c) What is the probability a light bulb will still function after 2,000 hours of use?

### Solution:

The lifetime of light bulbs follows an exponential distribution with a hazard rate of 0.001 failures per hour of use. Let $T$ denote the lifetime of light bulbs. For an exponential distribution, the cumulative hazard function for the hazard rate of $\lambda = 0.001$ is given by

$$H(t) = \int_{0}^t h(u)du = \int_{0}^t \lambda t|_{0}^t = \lambda t = 0.001 t$$
And the survival function is given by

$$S(t) = e^{-H(t)} = e^{-\lambda t} = e^{-0.001 t}$$
The probability density function is given by

$$f(t) = h(t) S(t) = \lambda e^{-\lambda t} = 0.001 e^{-0.001 t}$$
(a) The mean of the exponential random variable is given by

$$E(T) = \int_{0}^{\infty} S(t) dt = \int_{0}^{\infty} e^{-\lambda t} dt = 1/\lambda = 1/0.001 = 1000$$
Therefore, the mean lifetime of a randomly selected light bulb is 1,000 hours.



(b) The median of the exponential random variable satisfies $0.5 = e^{-\lambda t}$ so that is given by

$$t_{med} = \frac{\log(2)}{\lambda} = \frac{\log(2)}{0.001} = 1000 \log(2) = 301.03$$
That is, the median of a randomly selected light bulb is 301.03 hours.



(c) The probability a light bulb will still function after 2,000 hours of use is given by the probability density function

$$S(2000) = e^{-0.001 \times 2000} = 0.135$$

$$\\[1in]$$

## Question 2

The log-logistic distribution arises in a similar manner to the log-normal distribution. We say that $T \sim log-logistic(\mu, \sigma)$ if $Y = \ln T ∼ logistic(\mu, \sigma)$, where $\mu$ is the location parameter and $\sigma$ is the scale parameter of the logistic distribution. The probability density function of the logistic distribution is:

$$f(y) = \dfrac{\exp\bigg(\frac{y - \mu}{\sigma}\bigg)}{\sigma \Big[1 + \exp\bigg(\frac{y - \mu}{\sigma}\bigg)\Big]^2}$$
A commonly used reparametrization is $\alpha = 1/\sigma >0$ and $\lambda = \exp(-\frac{\mu}{\sigma})$.

(a) Derive an expression for the survival function of T using the parametrization in $\alpha$ and $\lambda$.
(b) Derive an expression for the hazard function of T using the parametrization in $\alpha$ and $\lambda$.
(c) The time to death (in days) following a kidney transplant follows a log-logistic distribution with $\alpha = 1.5$ and $\lambda = 0.01$.

i. Find the 50, 100, 150 day survival probabilities for kidney transplant patients.
ii. Find the median time to death following a kidney transplant.
iii. Show that the hazard rate is initially increasing and then decreasing over time. Find the time at which the hazard rate changes from increasing to decreasing.
iv. Find the mean time to death. $E(T) = \frac{\pi}{\sin (\pi/\alpha) \alpha \lambda^{1/\alpha}}$

### Solution:

(a) 


The reparametrized probability density function for logistic distribution is given as 

$$ f(y) = \dfrac{\exp\bigg(\frac{y - \mu}{\sigma}\bigg)}{\sigma \Big[1 + \exp\bigg(\frac{y - \mu}{\sigma}\bigg)\Big]^2} = \dfrac{\lambda \exp(\alpha y)}{\frac{1}{\alpha} \Big[ 1 + \lambda \exp(\alpha y )\Big]^2} = \dfrac{\alpha \lambda \exp(\alpha y)}{\Big[  1 + \lambda \exp(\alpha y )\Big]^2}$$
$$Y = \ln T \sim logistic(\alpha, \lambda) \implies T = \exp(Y) \sim log-logistic(\alpha, \lambda)$$

Writing $T$ as a function of $Y$ gives

$$T = g(Y) = \exp(Y) \implies Y = g^{-1}(T) = \ln(T)$$
Because the derivative of $\exp(Y)$ is always positive and g(Y) is strictly increasing, for the relevant support




$$f_T(t) = f_Y(g^{-1}(t)) \frac{dg^{-1}(t)}{dt} = \frac{\alpha \lambda \exp(\alpha g^{-1}(t))}{\bigg[1 +  \lambda \exp(\alpha g^{-1}(t))\bigg]^2}\cdot \frac{dg^{-1}(t)}{dt} = \frac{\alpha \lambda \exp(\alpha \ln(t))}{\bigg[1 +  \lambda \exp(\alpha \ln(t))\bigg]^2}\cdot \frac{d\ln(t)}{dt}$$

$$f_T(t) = \frac{\alpha \lambda \exp( \ln(t^{\alpha}))}{\bigg[1 + \lambda \exp(\ln(t^{\alpha}))\bigg]^2}\cdot \frac{1}{t} = \frac{\alpha \lambda t^{\alpha-1}}{\bigg[1 + \lambda t^{\alpha}\bigg]^2}$$

The cumulative distribution function (pdf) on support of $y$ could be represented as 

$$ F(t) = P(T \leq t) = \int_{-\infty}^t f(y) dt = \int_{-\infty}^t \frac{\alpha \lambda y^{\alpha-1}}{\bigg[1 + \lambda y^{\alpha}\bigg]^2}dy = \frac{\lambda t^{\alpha}}{1+\lambda t^{\alpha}}$$

Therefore, the survival function is given by 
$$S(t) = P (T >t) = 1 - F(t) = 1  -\frac{\lambda t^{\alpha}}{1+\lambda t^{\alpha}} = \frac{1}{1+\lambda t^{\alpha}}  $$



(b) The hazard function of T using the parametrization in $\alpha$ and $\lambda$ is given by

$$h(t) = \frac{f(t)}{S(t)} = \dfrac{\frac{\alpha \lambda t^{\alpha-1}}{\bigg[1 + \lambda t^{\alpha}\bigg]^2}}{\frac{1}{1+\lambda t^{\alpha}}}  = \frac{\alpha \lambda t^{\alpha-1}}{1+\lambda t^{\alpha}}$$
This form of $h(t)$ could also be derived using the expression $h(t) = -\frac{d}{dt}\ln S(t)$.

(c) The time to death (in days) following a kidney transplant follows a log-logistic distribution
with $\alpha = 1.5$ and $\lambda = 0.01$.


i. 
```{r}
kidney_survival = function(t){1/(1+(0.01*t^(1.5)))}

kidney_survival(50)
kidney_survival(100)
kidney_survival(150)
```



For $t = 50$, $S(t) = 0.2204812$. For $t = 50$, $S(t) = 0.09090909$. For $t = 50$, $S(t) = 0.0516231$.

Therefore, the 50, 100, and 150 day survival probabilities for kidney transplant patients is 22.04%, 9.09%, and 5.16%, respectively.


ii. Since the survival probability $S(t) = $, we find the median time by solving

$$0.5 = S(t) \implies \frac{1}{1+0.01 \times t^{1.5}} = 0.5 \implies t = (100)^{1/1.5} = 21.54435$$

iii. With $\lambda =0.01$ and $\alpha =1.5$, the hazard function becomes

$$h(t) = \frac{\alpha \lambda t^{\alpha-1}}{1+\lambda t^{\alpha}} = \frac{0.015t^{0.5}}{1 + 0.01t^{1.5}}$$
The first derivative of the hazard rate with respect to $t$ gives

$$h'(t) = \dfrac{(1 + 0.01t^{1.5})\cdot 0.0075t^{-0.5} - 0.015t^{0.5} \cdot 0.015 \cdot t^{0.5}}{(1 + 0.01t^{1.5})^2} =\frac{ 0.0075t^{-0.5} + 0.000075t - 0.000225t}{(1 + 0.01t^{1.5})^2} = 0.0075t^{-0.5} + 0.0003t  \implies t = 13.572$$
The following plot shows that the hazard rate is initially increasing and then decreasing over time.

```{r}
t=seq(0,100,0.1)
kidney_survival = function(t){((0.015*t^(0.5))/(1+(0.01*t^(1.5))))}
plot(t,kidney_survival(t),type='l',ylab="Hazard Rate")
```




iv. The mean time to death could be obtatined by solving

$$E(T) = \int_{0}^{\infty} t f(t) dt = \int_{0}^{\infty} t \frac{\alpha \lambda t^{\alpha-1}}{\bigg[1 + \lambda t^{\alpha}\bigg]^2}dt = \int_{0}^{\infty}\frac{\alpha \lambda t^{\alpha}}{\bigg[1 + \lambda t^{\alpha}\bigg]^2} dt$$

which solves to


$$E(T) =  \frac{\pi}{\sin (\pi/\alpha) \alpha \lambda^{1/\alpha}}$$

```{r}
#Mean Time to Death
Mean_time_death = pi/(sin(pi/1.5)*1.5*(0.01)^(1/1.5))
Mean_time_death
```
That is, the mean time to death is 52.10 days.


$$\\[1in]$$



## Question 3

Let $T$ denote the time to failure of an electrical component. The distribution of $T$ is exponential with MEAN $\theta/a$, where $a$ is the temperature at which the component is operated. Suppose that $n$ components are tested independently at KNOWN temperatures $a_1, a_2, \cdots, a_n$, respectively, and their observed lifetimes are $t_1, t_2, \cdots, t_n$. Note that the temperatures $a_1, a_2, \cdots, a_n$ are considered to be constants.
(a) Derive an expression for the maximum likelihood estimator (MLE) of $\theta$.
(b) Derive an expression for an estimate of the asymptotic variance of the MLE.


### Solution:

(a) The density function for the exponential distribution is given by

$$abc $$


The likelihood function is given by

$$\mathcal{L} = \prod_{i = 1}^n f(t) = \prod_{i = 1}^n \frac{a}{\theta} \exp(-at_i/\theta) = \bigg( \frac{a}{\theta}\bigg)^n \exp\big(-a/\theta\sum_{i=1}^n t_i\big)$$
And the log-likelihood is given by

$$\ell = \log(\mathcal{L}) = \log \bigg[ \prod_{i = 1}^n f(t) \bigg] = n \cdot \ln\bigg(\frac{a}{\theta} \bigg) - \frac{a}{\theta} \sum_{i=1}^n t_i$$
Taking the derivative of $\ell$ with respect to $ \theta$ gives

$$\frac{\partial \ell }{\partial \theta} = n \frac{\theta}{a} \bigg( -\frac{a}{\theta^2} \bigg) \quad - \quad  \sum_{i=1}^n t_i \bigg( -\frac{a}{\theta^2} \bigg)$$
Solving for optimality gives

$$ \hat{\theta}_{MLE} = a\frac{\sum_{i=1}^n t_i}{n} = a \bar{t}  $$



(b) Given that $T \sim \exp(mean(\frac{\theta}{a}))$, the expected value and variance are given by

$$ E(t) = \frac{\theta}{a} \quad and \quad V(t) = \frac{\theta^2}{a^2}$$
Also,
$$ E(\hat{\theta}_{MLE}) = \frac{a}{n}\times n \times \frac{\theta}{a} = \theta$$
$$E(\hat{\theta}_{MLE}^2) = E(a^2 \bar{t}^2) = a^2 E(\bar{t}^2) =  a^2 \bigg( V(\bar{t}) + E(\bar{t})^2\bigg) = a^2 \bigg( \dfrac{n\frac{\theta^2}{a^2}}{n^2} + (\frac{\theta}{a})^2\bigg) = \theta^2\frac{n+1}{n}$$

The expression for an estimate of the asymptotic variance of the MLE is given by

$$Var(\hat{\theta}_{MLE}) =  - E(\hat{\theta}_{MLE}^2) - E(\hat{\theta}_{MLE})^2  =\theta^2\frac{n+1}{n} - \theta^2 =  \frac{\theta^2}{n}$$

Asymptotically, as $n \rightarrow \infty$, then the variance converges to 0.


$$\\[1in]$$


## Question 4

Exercise 2.5 on page 24 of the text, Applied Survival Analysis Using R.
Consider a survival distribution with constant hazard $\lambda = 0.07$from $t = 0$ until $t=5$ and then hazard $\lambda = 0.14$from $t = 0$ for $t>5$ (This is known as a piecewise constant hazard.) Plot this hazard function and the corresponding survival function for $0 < t < 10$.What is the median survival time?


### Solution:


The survival function is given by $S(t) = \exp(-\int_0^t h(u) du$. That is, for $0 \geq t \geq 5$, $S(t) = e^{-0.07t}$. For $5 < t <10$, $S(t) = e^{-\int_0^5 0.07 du - \int_5^t 0.14 du} = e^{0.35 - 0.14(t-5)}$. 


```{r}

ht <- function(t){
    f <- NULL
    f[t>=0 & t<=5] <- 0.07
    f[t>5 ] <- 0.14
    f
}

t <- seq(0, 20, 0.1)
plot(t, ht(t), type="l", las=1, ylim=c(0,0.20), ylab="Hazard Rate", main="Plot of Hazard Function")
```

```{r}
# Define Hazard and Survival Function
hazard_function = function(t){
 ifelse(t <= 5, 0.07, 0.14)}
h_value = sapply(t,hazard_function)
S <- exp(-cumsum(h_value*0.001))
# Suvival funciton graph
plot(t,S,type="l",xlab="t",ylab="S(t)",main="Plot of the Survival Function")
```
$$\\[1in]$$


## Question 5

A model for lifetimes, with a bathtub-shaped hazard rate, is the distribution with survival function 
$$S(t) = \exp \{1 − \exp [(\lambda t)^\alpha]\} , \alpha, \lambda > 0, t > 0.$$

(a) If $\alpha =0.5$, show that the hazard rate has a bathtub shape and find the time at which
the hazard rate changes from decreasing to increasing.
(b) If $\alpha = 2$, show that the hazard rate of $t$ is monotone increasing.

### Solution:

(a) With survival function given by
$$S(t) = \exp \{1 − \exp [(\lambda t)^\alpha]\} , \alpha, \lambda > 0, t > 0.$$
The cumulative distirbution function is given by

$$F(t) = 1- S(t) = 1 -  \exp \{1 − \exp [(\lambda t)^\alpha]\} $$
The probability density function is given by

$$f(t)= \frac{dF(t)}{dt} = -\exp \{1 − \exp [(\lambda t)^\alpha]\}  \cdot − \exp [(\lambda t)^\alpha] \cdot \alpha (\lambda t)^{\alpha - 1} \lambda = \alpha \lambda (\lambda t)^{\alpha - 1}\cdot \exp [(\lambda t)^\alpha] \cdot \exp \{1 − \exp [(\lambda t)^\alpha]\}$$
Therefore, the hazard rate is given by

$$h(t) = \frac{f(t)}{S(t)} = \alpha \lambda (\lambda t)^{\alpha - 1}\cdot \exp [(\lambda t)^\alpha]$$
If $\alpha = 0.5$, then the hazard rate is

$$h(t) = 0.5 \lambda (\lambda t)^{-0.5}\cdot \exp [(\lambda t)^{0.5}]$$


Assume $\lambda = 1$, then the plot appears as below:

```{r}
t=seq(0,4,0.1)
y=0.5*1*(1*t)^(0.5-1)*exp((1*t)^(0.5))
plot(t,y, main="Plot for alpha = 0.5 and lambda = 1", type='l')
```

Assume $\lambda = 2$, then the plot appears as below:

```{r}
t=seq(0,4,0.1)
y=0.5*2*(2*t)^(0.5-1)*exp((2*t)^(0.5))
plot(t,y, main="Plot for alpha = 0.5 and lambda = 2", type='l')
```
$$h'(t) = \frac{d h(t)}{dt} = \frac{d }{dt}\bigg( \alpha \lambda (\lambda t)^{\alpha - 1}\cdot \exp [(\lambda t)^\alpha]\bigg)=\frac{d }{dt}\bigg( \alpha \lambda^{\alpha} (t)^{\alpha - 1}\cdot \exp [(\lambda t)^\alpha]\bigg)\\
=\alpha \lambda^{\alpha}\bigg( (t)^{\alpha - 1}\cdot \frac{d }{dt}\exp [(\lambda t)^\alpha]+\exp [(\lambda t)^\alpha]\frac{dt }{dt}\bigg)\\
=\alpha \lambda^{\alpha}\bigg( (t)^{\alpha - 1}\cdot \exp [(\lambda t)^\alpha]\cdot \alpha(\lambda t)^{\alpha -1} \cdot \lambda+\exp [(\lambda t)^\alpha]\bigg)\\
=\alpha \lambda^{\alpha}\exp [(\lambda t)^\alpha]\bigg( \alpha \lambda^{\alpha}(t)^{2(\alpha - 1)} +1\bigg)$$

Solving $h'(t) = 0$ gives the time at which the hazard rate changes from decreasing to increasing

$$t = \bigg(-\frac{1}{2\lambda^{\alpha}}\bigg)^{\frac{1}{2(\alpha-1)}}$$




(b) If $\alpha = 2$, the first derivative of the hazard function changes to


$$2 \lambda^{2}\exp [(\lambda t)^2]\bigg( 2 \lambda^{2}(t)^{2} +1\bigg) $$

Since $\lambda >0$ and any positive $t$, $t >0$,implies that this function is essentially a constant and two increasing functions multiplied together, which implies that it is strictly positive. That is, the hazard function is strictly increasing, or monotone increasing. 


